---
title: "00_Setup_and_preprocessing"
author: 
- name: "Ralitsa R. Madsen"
  affiliation: UCL
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: true
  html_document:
    code_folding: show
    theme: paper
    highlight: kate
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
---

# Biological background 

Experiment perfmored by Elena Guidamillas, postdoctoral researcher in Bart Vanhaesebroeck's group at UCL, London. 

For conditions, see metadata and the accompanying readme file.

Focus of this notebook: analysis of U87MG glioblastoma cancer cells - WT or PIK3CD KO; +/- 2 different PI3Kd inhibitors.


#### Analysis 


Pipeline for further differential gene expression analysis in R: 
  - Standard limma-voom workflow analysis
  - Relevant references (doi): 10.1093/bioinformatics/btt656 ; doi: 10.1093/nar/gkv007 


# Setup

```{r setup, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F) # setting global chunk options for knitted output
#source("scripts/install.packages.R") # installs all required packages; uncomment and run if not done once already
source("scripts/packages.R") # loads the required packages
sessionInfo() # package version information
input.data <- file.path(getwd(), "data_input/")
output.figures <- file.path(getwd(), "figures_output/")
output.df <- file.path(getwd(), "data_output/")
options("scipen"=5, "digits"=4) # control scientific notation, in this case triggered only when number is 10 digits
```

# Metadata

```{r}
metadata <- read_csv(paste0(input.data, "metadata_ELG.csv")) %>% 
  mutate(TREATMENT = factor(TREATMENT, 
                          levels = c("vehicle", "inh1", "inh2")),
         REPLICATE = factor(REPLICATE)
         ) %>% 
  filter(CELL_LINE == "U87MG") # ensuring focus on U87MG only
metadata <- as.data.frame(metadata)
```

```{r}
library(assertr)
metadata %>% 
  assert_rows(col_concat, is_uniq, FOLDER) # confirming that the FOLDER IDs are unique for all rows - necessary for downstream sample assignment
```


- Check factorial levels and replicate numbers

```{r}
metadata %>% 
  group_by(GENOTYPE, CELL_LINE, TREATMENT) %>% 
  dplyr::summarise(n = n()) # all collrect n=3
```
- Create new column with meaningful sample name 

```{r}
metadata <- metadata %>% 
  unite("Rcode_analysis", CELL_LINE, GENOTYPE, TREATMENT, REPLICATE, remove = FALSE)
rownames(metadata) <- metadata$FOLDER # required for some of the downstream functions to run correctly 
metadata #check 
```


# Read data & merge with metadata

```{r raw_data}
rawGeneCountTable <- read.table(paste0(input.data, 
                                       "merged_gene_counts.txt"), 
                                header = TRUE,
                            row.names = 1) # read raw gene count data file, columns correspond to counts in individual samples, rows to genes
###
ncol(rawGeneCountTable) # 45 columns = correct
```


```{r}
head(rawGeneCountTable) # inspect
```

```{r}
#select only U87MG samples
rawGeneCountTable <- rawGeneCountTable[, colnames(rawGeneCountTable) %in% rownames(metadata)]
ncol(rawGeneCountTable) # 18 is correct
```



```{r annotate}
index <- match(colnames(rawGeneCountTable), rownames(metadata)) # ordering sample metadata information according to column names of raw data table; stored as index values for subsequent subsetting
names.samples <- as.factor(metadata$Rcode_analysis[index])
exp.samples <- as.factor(metadata$REPLICATE[index])
treatment.samples <- as.factor(metadata$TREATMENT[index])
genotype.samples <- as.factor(metadata$GENOTYPE[index])
cell.samples <- as.factor(metadata$CELL_LINE[index])
colnames(rawGeneCountTable) <- names.samples
```

```{r}
head(rawGeneCountTable) # confirm that this worked
```

```{r}
rawGeneCountTable.DGE <-
  DGEList(rawGeneCountTable) # create DGEobject, necessary for downstream analysis using edgeR functions
# This function will create a combined table of counts with rows for genes and columns for samples.
rawGeneCountTable.DGE$samples$treatment <- treatment.samples
rawGeneCountTable.DGE$samples$exp <- exp.samples
rawGeneCountTable.DGE$samples$genotype <- genotype.samples
rawGeneCountTable.DGE$samples$cell.samples <- cell.samples
rawGeneCountTable.DGE$samples # check that annotations are correct
```

- Check range and mean of library size

```{r}
range(rawGeneCountTable.DGE[[2]][2]) # very large range; from ~21M to ~60M
```

```{r}
mean(rawGeneCountTable.DGE$samples$lib.size) # mean library size: [1] c. 30M
median(rawGeneCountTable.DGE$samples$lib.size) # mean library size: [1] c. 25M
```

- How many genes are not expressed at all across all samples, i.e. counts = 0.

```{r}
table(rowSums(rawGeneCountTable.DGE$counts == 0) == 18)
```

## Filtering

Filtering out lowly expressed genes using quantile method from TCGAbiolinks package

  - Determine reasonable filtering threshold based on the below loop results

```{r}
for (value in c(0, 0.1, 0.25, 0.3, 0.4, 0.5, 0.6, 0.7, 0.75, 0.8, 0.9)) {
  dataFilt <- TCGAanalyze_Filtering(rawGeneCountTable.DGE$counts,
                                    method = "quantile", 
                                    qnt.cut =  value) # looping through possible quantile values
  lcpm.dataFilt <- log2(dataFilt+0.5) # not actually transformed to cpm prior to this, so seeing effect of 2 outlier samples with respect to library size.
  nsamples1 <- ncol(lcpm.dataFilt) 
  plot(
    density(lcpm.dataFilt[, 1]), 
    lwd = 2,
    ylim = c(0, 0.5),
    las = 2,
    # axis labels perpendicular to the axes; lwd = line width
    main = "",
    xlab = ""
  )
  abline(v = 0, lty = 3) # lty = line type, dotted in this case
  # Nested loop to plot density of each sample following filtering 
  for (i in 2:nsamples1) {
    den <- density(lcpm.dataFilt[, i])
    lines(den$x, den$y, lwd = 2)
  }
  title(main = paste0("Raw_data, quantile=", value), xlab = "logged counts")
}
```


- A cut-off of 0.7 appears reasonable

```{r}
cutoff.filt <- 0.7
```


```{r}
rawGeneCountTable.DGE$counts.filt <- TCGAanalyze_Filtering(rawGeneCountTable.DGE$counts,
                                                           method = "quantile", 
                                                           qnt.cut =  cutoff.filt) # down to 17636 genes

dim(rawGeneCountTable.DGE$counts.filt) # 17636 genes retained
```



- Create new ExpressionSet object of filtered counts

```{r}
rawGeneCountTable.DGE.filt <- DGEList(rawGeneCountTable.DGE$counts.filt)
```

## MDS plot


### MDS plot before normalisation

- Colour according to different grouping criteria to check for drivers of separation


```{r}
col.exp <- exp.samples
levels(col.exp) <- brewer.pal(length(levels(exp.samples)), "Set1") 

col.genotype <- genotype.samples
levels(col.genotype) <- brewer.pal(length(levels(genotype.samples)), "Set1") 

col.treatment <- treatment.samples
levels(col.treatment) <- brewer.pal(length(levels(col.treatment)), "Set1") 

col.cell <- cell.samples
levels(col.cell) <- brewer.pal(length(levels(cell.samples)), "Set1") 
```


 - Colour according to genotype

 
```{r}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "MDS_EG_pre.norm_U87MG_genotype.col_500genes", ".png"), 
    height=1500, width=2000, res=300)
par(mar=c(4.1, 4.1, 4.1, 4.1), xpd=TRUE) # add extra space to the right of the plot; needed to include legend and still see samples
plotMDS(
    cpm(rawGeneCountTable.DGE$counts.filt, log = T),
    labels = as.character(genotype.samples),
    pch = 1,
    col = as.character(col.genotype),
    xlim = c(-3, 3),
    ylim = c(-2, 2),
    xlab = "Dimension 1",
    ylab = "Dimension 2",
    cex = 1,
    cex.axis = 1, 
    cex.lab = 1
  )
title(main = "MDS_EG_U87_PI3Kd MDS plot \n Leading fold-change (500 most divergent genes)", cex.main=1)
legend(
  "topright",
  fill = paste0(levels(col.genotype)),
  legend = paste0(levels(genotype.samples)),
  cex = 0.8
)
dev.off()
```


 - Colour according to treatment

 
```{r}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "MDS_EG_pre.norm_U87MG_treatment.col_500genes", ".png"), 
    height=1500, width=2000, res=300)
par(mar=c(4.1, 4.1, 4.1, 4.1), xpd=TRUE) # add extra space to the right of the plot; needed to include legend and still see samples
plotMDS(
    cpm(rawGeneCountTable.DGE$counts.filt, log = T),
    labels = as.character(treatment.samples),
    pch = 1,
    col = as.character(col.genotype),
    xlim = c(-3, 3),
    ylim = c(-2, 2),
    xlab = "Dimension 1",
    ylab = "Dimension 2",
    cex = 1,
    cex.axis = 1, 
    cex.lab = 1
  )
title(main = "MDS_EG_U87_PI3Kd MDS plot \n Leading fold-change (500 most divergent genes)", cex.main=1)
legend(
  "topright",
  fill = paste0(levels(col.genotype)),
  legend = paste0(levels(genotype.samples)),
  cex = 0.8
)
dev.off()
```

Inh1 contributes to separation in 2nd dimension. 


## TMM normalisation

- Reference for method: Robinson, M.D., Oshlack, A., 2010. A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biol. 11, R25. https://doi.org/10.1186/gb-2010-11-3-r25

```{r}
rawGeneCountTable.DGE.filt <- calcNormFactors(rawGeneCountTable.DGE.filt, 
                                              method = "TMM")
```

```{r}
rawGeneCountTable.DGE.filt$samples # check normalisation factors
```

### MDS plot post-normalisation

  - Check how the samples cluster following normalisation; for this, first run voom so that normalisation factors can be taken into account.

```{r normalisation_voom}
voom_normalisation <- voom(rawGeneCountTable.DGE.filt) 
```


```{r}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "MDS_EG_post.norm_cell_line_treatment.col_500genes", ".png"), 
    height=1500, width=2000, res=300)
par(mar=c(4.1, 4.1, 4.1, 4.1), xpd=TRUE) # add extra space to the right of the plot; needed to include legend and still see samples
plotMDS(
    cpm((2^voom_normalisation$E), log=T),
    labels = as.character(treatment.samples),
    pch = 1,
    col = as.character(col.genotype),
    xlim = c(-3, 3),
    ylim = c(-2, 2),
    xlab = "Dimension 1",
    ylab = "Dimension 2",
    cex = 1,
    cex.axis = 1, 
    cex.lab = 1
  )
title(main = "MDS_EG_U87_PI3Kd MDS plot \n Leading fold-change (500 most divergent genes)", cex.main=1)
legend(
  "topright",
  fill = paste0(levels(col.genotype)),
  legend = paste0(levels(genotype.samples)),
  cex = 0.8
)
dev.off()
```

### PCA pre-normalisation

For additional information & functions within PCAtools package, see: https://bioconductor.org/packages/release/bioc/vignettes/PCAtools/inst/doc/PCAtools.html


```{r}
# check that sample names match exactly between pdata and expression data 
rownames(metadata) <- metadata$Rcode_analysis
all(colnames(voom_normalisation) == rownames(metadata))
```


```{r}
p <- pca(cpm(rawGeneCountTable.DGE.filt, log = T), 
         metadata = metadata, 
         removeVar = 0.1) ## -- removing the lower 10% of variables based on variance
```


#### Scree plot

```{r screeplot}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "PCA_screeplot_U87MG", ".png"), 
    height=1500, width=2000, res=300)
screeplot(p, axisLabSize = 10, titleLabSize = 12)
dev.off()
```

- Which PCs contribute 80% of the variance in the data?

```{r}
which(cumsum(p$variance) > 80)[1] # the first 5 
```

#### Bi-plot

- PC1 vs PC2

```{r}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "PC1vsPC2_U87MG", ".png"), 
    height=1500, width=2000, res=300)
biplot(p, x = "PC1", y = "PC2",
  lab = NULL,
  colby = "TREATMENT",
  shape = "GENOTYPE",
  hline = 0, vline = 0,
      # ellipse config
  ellipse = F,
  ellipseConf = 0.95,
  ellipseFill = TRUE,
  ellipseAlpha = 1/4,
  ellipseLineSize = 1.0,
  gridlines.major = FALSE, gridlines.minor = FALSE,
  title = "PCA bi-plot",
  subtitle = "PC1 versus PC2",
  legendPosition = "right")
dev.off()
```


- PC1 vs PC3


```{r}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "PC1vsPC3_U87MG", ".png"), 
    height=1500, width=2000, res=300)
biplot(p, x = "PC1", y = "PC3",
  lab = NULL,
  colby = "TREATMENT",
  shape = "GENOTYPE",
  hline = 0, vline = 0,
      # ellipse config
  ellipse = F,
  ellipseConf = 0.95,
  ellipseFill = TRUE,
  ellipseAlpha = 1/4,
  ellipseLineSize = 1.0,
  gridlines.major = FALSE, gridlines.minor = FALSE,
  title = "PCA bi-plot",
  subtitle = "PC1 versus PC3",
  legendPosition = "right")
dev.off()
```

- PC2 vs PC3

```{r}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "PC2vsPC3_U87MG", ".png"), 
    height=1500, width=2000, res=300)
biplot(p, x = "PC2", y = "PC3",
  lab = NULL,
  colby = "TREATMENT",
  shape = "GENOTYPE",
  hline = 0, vline = 0,
      # ellipse config
  ellipse = F,
  ellipseConf = 0.95,
  ellipseFill = TRUE,
  ellipseAlpha = 1/4,
  ellipseLineSize = 1.0,
  gridlines.major = FALSE, gridlines.minor = FALSE,
  title = "PCA bi-plot",
  subtitle = "PC2 versus PC3",
  legendPosition = "right")
dev.off()
```

- PC1 vs PC4;

```{r}
png(file=paste0(output.figures, 
                Sys.Date(), 
                sep="_", "PC1vsPC4_U87MG", ".png"), 
    height=1500, width=2000, res=300)
biplot(p, x = "PC1", y = "PC4",
  lab = NULL,
  colby = "TREATMENT",
  shape = "GENOTYPE",
  hline = 0, vline = 0,
      # ellipse config
  ellipse = F,
  ellipseConf = 0.95,
  ellipseFill = TRUE,
  ellipseAlpha = 1/4,
  ellipseLineSize = 1.0,
  gridlines.major = FALSE, gridlines.minor = FALSE,
  title = "PCA bi-plot",
  subtitle = "PC1 versus PC4",
  legendPosition = "right")
dev.off()
```


******

Proceed with RNotebook "01_Differential_gene_expression_analysis.Rmd".
