---
title: "01_Differential_gene_expression_analysis"
author: 
- name: "Ralitsa R. Madsen"
  affiliation: UCL
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: true
  html_document:
    code_folding: show
    theme: paper
    highlight: kate
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
---

# Readme

This notebook should only be run after "00_Setup_and_preprocessing.Rmd" has been run, with access to the same Global Environment. 

The differential gene expression analysis performed within this RNotebook is based on the voom/limma/ebayes workflow, as outlined in the following references: 

  - Smyth et al. 2016 limma: linear models for microarray and RNA-seq data user's guide; pp. 43-45 (or full Chapter 9)
  - https://support.bioconductor.org/p/93742/


# Setup

```{r setup, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F) # setting global chunk options for knitted output
#source("scripts/install.packages.R") # installs all required packages; uncomment and run if not done once already
source("scripts/packages.R") # loads the required packages
sessionInfo() # package version information
input.data <- file.path(getwd(), "data_input/")
output.figures <- file.path(getwd(), "figures_output/")
output.df <- file.path(getwd(), "data_output/")
options("scipen"=5, "digits"=4) # control scientific notation, in this case triggered only when number is 10 digits
```

# Design & contrast matrix setup

```{r}
print(metadata)
```



```{r}
genotype.treatment <- paste(genotype.samples, treatment.samples, sep=".")
design.matrix <-
  model.matrix( ~ 0 + genotype.treatment + exp.samples) # note inclusion of replicate experiment to regress any batch effects though preprocessing suggested these to be minor
design.matrix #check
```


```{r}
contr.matrix <- makeContrasts(
  KO_INH1 =  genotype.treatmentKO.inh1-genotype.treatmentKO.vehicle,
  KO_INH2 =  genotype.treatmentKO.inh2-genotype.treatmentKO.vehicle,
  WT_INH1 =  genotype.treatmentWT.inh1-genotype.treatmentWT.vehicle,
  WT_INH2 =  genotype.treatmentWT.inh2-genotype.treatmentWT.vehicle,
  KOvsWT = ((genotype.treatmentKO.inh1+genotype.treatmentKO.inh2+genotype.treatmentKO.vehicle)/3) - ((genotype.treatmentWT.inh1+genotype.treatmentWT.inh2+genotype.treatmentWT.vehicle)/3),
  levels = colnames(design.matrix)
)
contr.matrix
```

# Voom, limma, eBayes 

- Model the mean-variance relations with the voom method; see doi: 10.1186/gb-2014-15-2-r29


```{r}
v.DGE <-
  voom(rawGeneCountTable.DGE.filt, design.matrix, plot = TRUE) 
```


```{r}
vfit.DGE <-
  lmFit(v.DGE, design.matrix) # linear modelling; 
# uses the normalised expression values on the log2 scale (calculated by voom)
# and the inverse variance weights (calculated by voom)
vfit.DGE <-
  contrasts.fit(vfit.DGE, contrasts = contr.matrix)
efit.DGE <-
  eBayes(vfit.DGE, proportion = 0.01) # default proportion = assumed proportion of genes that are differentially expressed
```


```{r}
plotSA(efit.DGE) # heteroscedascity removed 
```

# Differentially expressed genes

## Summary

- Control FDR (Benjamini-Hochberg FDR at 5 %). Using method "separate" in this case - "global" is more stringent as it effectively would the a multiple comparison across 5x ~18000 genes instead of 1x only = correcting for multiple comparisons within each contrast rather than globally. 

```{r}
summary(decideTests(efit.DGE, method = "separate")) 
```

- Test effect of 'magnitude' of differential expression by varying fold-change cut-off

```{r}
summary(decideTests(efit.DGE, method = "separate", lfc=log2(2))) 
```


```{r}
summary(decideTests(efit.DGE, method = "separate", lfc=log2(1.3))) 
```



## BioMart annotation & data export

Export the limma object and read again as a dataframe for easier access to the results. The following variables are extracted: 

1. The average log2-intensity (A)
2. The coefficients or contrasts (log2-fold-changes) 
3. Moderated t-statistics 
4. t-statistic P-values (including B-H adjusted p-values)
5. F-statistic if available
6. F-statistic P-values if available
7. Classification if available
8. Gene names and annotation.

```{r max.print = 15000}
write.fit(efit.DGE,
          file = paste0(output.df, Sys.Date(), sep="_", "limmma_voom_DE_U87_PI3Kd_ELG_BV", ".txt"),
          adjust = "BH",
          method = "separate") # Note that you don't specify lfc in this case because it outputs all genes with the adjusted p-value 
de.data <- 
  read.delim(paste0(output.df, Sys.Date(), sep="_", "limmma_voom_DE_U87_PI3Kd_ELG_BV", ".txt"),
             sep = "",
             header = T)
```

```{r}
de.data$Genes <- rownames(de.data)
de.data
```

- Interface with BioMart to retrieve gene annotation

```{r}
gene.mart.gene <-
  getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol", "entrezgene_id", "uniprotswissprot"), # unitprot and HGNC needed for many downstream functions 
    filters = "ensembl_gene_id",
    values = de.data$Genes,
    mart = mart
  )
colnames(gene.mart.gene) <- c("Genes", "HGNC", "Entrez", "uniprot")
gene.mart.gene <-
  gene.mart.gene[!duplicated(gene.mart.gene$Genes), ] 
de.data.annot <-
  left_join(de.data, gene.mart.gene, by = "Genes")
##
de.data.annot # still 17636 columns
```

```{r}
# Check for PIK3CD
de.data.annot %>% 
  filter(grepl("PIK3CA|PIK3CD|PIK3CB", HGNC))
```

- Extract background gene set for external pathway analyses

```{r}
write.table(de.data.annot %>% 
              filter(Genes != "") %>% 
              .$Genes, 
            file = paste0(output.df, Sys.Date(), sep="_", "bgd_genes_ENSG_U87_PI3KD", ".txt"), row.names=F)
```

```{r}
write.table(de.data.annot %>% 
              filter(HGNC != "") %>% 
              .$HGNC, 
            file = paste0(output.df, Sys.Date(), sep="_", "bgd_genes_HGNC_U87_PI3KD", ".txt"), row.names=F)
```


- Write out the annotated gene expression analysis results

```{r}
write_csv(de.data.annot,
          file=paste0(output.df, Sys.Date(), sep="_", "limmma_voom_DE_U87_PI3Kd_ELG_BV_annotated.csv"))
```



- Write out the log2 expression values from voom, alongside gene annotation


```{r}
voom_values <- as.data.frame(v.DGE$E)
voom_values$Genes <- rownames(voom_values)
```

```{r}
voom_values_annot <-
  left_join(voom_values, gene.mart.gene, by = "Genes")
```

```{r}
voom_values_annot %>% 
  filter(grepl("PIK3CA|PIK3CD|PIK3CB", HGNC))
```


```{r}
write_csv(voom_values_annot,
          file=paste0(output.df, Sys.Date(), sep="_", "voom_values_U87_PIK3d_ELG_BV_annotated.csv"))
```


```{r}
results <- decideTests(efit.DGE, adjust.method = "BH", 
                       method = "separate",
                       lfc=log2(1.3))
results.df <- as.data.frame(results)
results.df$Genes <- rownames(results)
```

- Retrieve HGNC annotation for above 

```{r}
gene.mart.gene.1 <-
  getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = results.df$Genes,
    mart = mart
  )
colnames(gene.mart.gene.1) <- c("Genes", "HGNC")
gene.mart.gene.1 <-
  gene.mart.gene.1[!duplicated(gene.mart.gene.1$Genes), ] 
results.df <-
  left_join(results.df, gene.mart.gene.1, by = "Genes")
###
results.df #17,636 rows = correct
```
******

Proceed with RNotebook "02_GSEA_pathway_analyses.Rmd".

******


