---
title: "02_GSEA_pathway_analyses"
author: 
- name: "Ralitsa R. Madsen"
  affiliation: UCL
output: 
  html_notebook:
    theme: united
    toc: yes
    toc_float: true
  html_document:
    code_folding: show
    theme: paper
    highlight: kate
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
---

# Readme

This notebook should only be run after "01_Differential_gene_expression_analysis.Rmd" has been run, with access to the same Global Environment. 

# Setup

```{r setup, warning=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F) # setting global chunk options for knitted output
#source("scripts/install.packages.R") # installs all required packages; uncomment and run if not done once already
source("scripts/packages.R") # loads the required packages
sessionInfo() # package version information
input.data <- file.path(getwd(), "data_input/")
output.figures <- file.path(getwd(), "figures_output/")
output.df <- file.path(getwd(), "data_output/")
options("scipen"=5, "digits"=4) # control scientific notation, in this case triggered only when number is 10 digits
```

# GSEA testing with fgsea

Reference: Sergushichev, A.A., 2016. An algorithm for fast preranked gene set enrichment analysis using cumulative statistic calculation. bioRxiv 060012.

#### Retrieve relevant gene sets

Hallmark gene sets from mSigDb

```{r}
m_df <- msigdbr(species = "Homo sapiens", category = "H")
```

```{r}
m_df1 <- m_df %>% dplyr::select(gs_name, human_gene_symbol)
```

```{r}
unique.gs.h <- unique(m_df1$gs_name)
length <- length(unique.gs.h)
hallmark.gs <- vector(mode = "list", length = length)
```

```{r}
m_df1[m_df1$gs_name %in% unique.gs.h[1],]
```

Create a list of the hallmark gene sets

```{r}
for (gs in 1:length) {
  gene.set.h <- unique.gs.h[gs] # pull out signature name
  names(hallmark.gs)[gs] <- gene.set.h # assign signature name to list components
  hallmark.gs[[gs]] <- m_df1$human_gene_symbol[m_df1$gs_name %in% gene.set.h] # assign signature gene ids to the respective list component
}
```

```{r}
length(hallmark.gs)
```


## Create vectors of gene expression values from the different lists of differentially expressed genes

NB: these vectors won't be ranked manually as this happens during the fgsea run itself.

- First make sure that any ENSGs resulting in duplicate HGNCs are removed to avoid ambiguity in the subsequent fgsea analysis

```{r}
length(de.data.annot$HGNC[duplicated(de.data.annot$HGNC)]) # 4120
```

- Confirm that these duplications do not reflect duplications at the ENSG level

```{r}
length(de.data.annot$Genes[duplicated(de.data.annot$Genes)]) # 0, i.e. no ENSG duplications
```

- Extract the duplicated genes

```{r}
dupl <- unique(de.data.annot$HGNC[duplicated(de.data.annot$HGNC)]) # actually only 1 real duplication = TBCE; the rest come from missing annotations 
de.data.annot.no.dupl <- de.data.annot %>% 
  filter(!(HGNC %in% dupl))
```


```{r}
# KOvsWT
de.data.annot.KOvsWT.rank <- de.data.annot.no.dupl %>% 
  dplyr::select(HGNC, t.KOvsWT) %>% 
  filter(HGNC != "") %>% # already removed, but included from external code snippet
  filter(!is.na(HGNC)) 
de.data.annot.KOvsWT.rank.vector <- tibble::deframe(de.data.annot.KOvsWT.rank)
```

```{r}
head(de.data.annot.KOvsWT.rank.vector)
```

```{r}
# WT INH1vsVEH
de.data.annot.WT_INH1vsVEH.rank <- de.data.annot.no.dupl %>% 
  dplyr::select(HGNC, t.WT_INH1) %>% 
  filter(HGNC != "") %>% # already removed, but included from external code snippet
  filter(!is.na(HGNC)) 
de.data.annot.WT_INH1vsVEH.rank.vector <- tibble::deframe(de.data.annot.WT_INH1vsVEH.rank)
```

```{r}
head(de.data.annot.WT_INH1vsVEH.rank.vector)
```

```{r}
# WT INH2vsVEH
de.data.annot.WT_INH2vsVEH.rank <- de.data.annot.no.dupl %>% 
  dplyr::select(HGNC, t.WT_INH2) %>% 
  filter(HGNC != "") %>% # already removed, but included from external code snippet
  filter(!is.na(HGNC)) 
de.data.annot.WT_INH2vsVEH.rank.vector <- tibble::deframe(de.data.annot.WT_INH2vsVEH.rank)
```

```{r}
head(de.data.annot.WT_INH2vsVEH.rank.vector)
```


```{r}
# KO INH1vsVEH
de.data.annot.KO_INH1vsVEH.rank <- de.data.annot.no.dupl %>% 
  dplyr::select(HGNC, t.KO_INH1) %>% 
  filter(HGNC != "") %>% # already removed, but included from external code snippet
  filter(!is.na(HGNC)) 
de.data.annot.KO_INH1vsVEH.rank.vector <- tibble::deframe(de.data.annot.KO_INH1vsVEH.rank)
```

```{r}
head(de.data.annot.KO_INH1vsVEH.rank.vector)
```

```{r}
# KO INH2vsVEH
de.data.annot.KO_INH2vsVEH.rank <- de.data.annot.no.dupl %>% 
  dplyr::select(HGNC, t.KO_INH2) %>% 
  filter(HGNC != "") %>% # already removed, but included from external code snippet
  filter(!is.na(HGNC)) 
de.data.annot.KO_INH2vsVEH.rank.vector <- tibble::deframe(de.data.annot.KO_INH2vsVEH.rank)
```

```{r}
head(de.data.annot.KO_INH2vsVEH.rank.vector)
```


### fgsea run 

Using updated approach where default is fgseaMultilevel

Not focusing on INH2 due to small t values, i.e. very few differentially expressed genes across the two genotypes.

##### KOvsWT

```{r}
fgsea.de.data.annot.KOvsWT <- fgsea(pathways = hallmark.gs, 
                  stats = de.data.annot.KOvsWT.rank.vector,
                  minSize=15,
                  maxSize=500)
```


```{r}
fgsea.de.data.annot.KOvsWT %>% 
  arrange(padj)
```

Make table plot

```{r}
topPathwaysUp.KOvsWT <- fgsea.de.data.annot.KOvsWT[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown.KOvsWT <- fgsea.de.data.annot.KOvsWT[ES < 0][head(order(pval), n=10), pathway]
topPathways.KOvsWT <- c(topPathwaysUp.KOvsWT, rev(topPathwaysDown.KOvsWT))
png(paste0(output.figures, Sys.Date(), sep="_", "GSEA_KOvsWT", ".png"),
    height=1500, width=3500, res=300)
plotGseaTable(hallmark.gs[topPathways.KOvsWT], 
              de.data.annot.KOvsWT.rank.vector, 
              fgsea.de.data.annot.KOvsWT, 
              gseaParam = 0.5)
dev.off()
```

```{r}
fgsea.de.data.annot.KOvsWT$Condition <- "KOvsWT" # required for heatmap
```


##### WT INH1vsVEH

```{r}
fgsea.de.data.annot.WT.INH1vsVEH <- fgsea(pathways = hallmark.gs, 
                  stats = de.data.annot.WT_INH1vsVEH.rank.vector,
                  minSize=15,
                  maxSize=500)
```


```{r}
fgsea.de.data.annot.WT.INH1vsVEH %>% 
  arrange(padj)
```

Make table plot

```{r}
topPathwaysUp.WT.INH1vsVEH <- fgsea.de.data.annot.WT.INH1vsVEH[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown.WT.INH1vsVEH <- fgsea.de.data.annot.WT.INH1vsVEH[ES < 0][head(order(pval), n=10), pathway]
topPathways.WT.INH1vsVEH <- c(topPathwaysUp.WT.INH1vsVEH, rev(topPathwaysDown.WT.INH1vsVEH))
png(paste0(output.figures, Sys.Date(), sep="_", "GSEA_WT_INH1vsVEH", ".png"),
    height=1500, width=3500, res=300)
plotGseaTable(hallmark.gs[topPathways.WT.INH1vsVEH], 
              de.data.annot.WT_INH1vsVEH.rank.vector, 
              fgsea.de.data.annot.WT.INH1vsVEH, 
              gseaParam = 0.5)
dev.off()
```

```{r}
fgsea.de.data.annot.WT.INH1vsVEH$Condition <- "WT.INH1vsVEH" # required for heatmap
```

##### KO INH1vsVEH

```{r}
fgsea.de.data.annot.KO.INH1vsVEH <- fgsea(pathways = hallmark.gs, 
                  stats = de.data.annot.KO_INH1vsVEH.rank.vector,
                  minSize=15,
                  maxSize=500)
```


```{r}
fgsea.de.data.annot.KO.INH1vsVEH %>% 
  arrange(padj)
```

Make table plot

```{r}
topPathwaysUp.KO.INH1vsVEH <- fgsea.de.data.annot.KO.INH1vsVEH[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown.KO.INH1vsVEH <- fgsea.de.data.annot.KO.INH1vsVEH[ES < 0][head(order(pval), n=10), pathway]
topPathways.KO.INH1vsVEH <- c(topPathwaysUp.KO.INH1vsVEH, rev(topPathwaysDown.KO.INH1vsVEH))
png(paste0(output.figures, Sys.Date(), sep="_", "GSEA_KO_INH1vsVEH", ".png"),
    height=1500, width=3500, res=300)
plotGseaTable(hallmark.gs[topPathways.KO.INH1vsVEH], 
              de.data.annot.KO_INH1vsVEH.rank.vector, 
              fgsea.de.data.annot.KO.INH1vsVEH, 
              gseaParam = 0.5)
dev.off()
```

```{r}
fgsea.de.data.annot.KO.INH1vsVEH$Condition <- "KO.INH1vsVEH" # required for heatmap
```



##### Heatmap

```{r}
hallmark.GSEA.wide <- rbind(fgsea.de.data.annot.KOvsWT %>% dplyr::select(-leadingEdge),
                            fgsea.de.data.annot.WT.INH1vsVEH %>% dplyr::select(-leadingEdge),
                            fgsea.de.data.annot.KO.INH1vsVEH %>% dplyr::select(-leadingEdge)) %>% 
  #filter(padj < 0.1) %>% # filter out for no p-value
  dplyr::select(pathway, NES, Condition) %>% 
  arrange(Condition, NES) %>% # doing this to ensure sensible ordering of heatmap further below
  pivot_wider(names_from = Condition, values_from = NES)
```


```{r}
head(hallmark.GSEA.wide)
```


```{r}
hallmark.GSEA.wide.m <- as.matrix(hallmark.GSEA.wide[,-1])
rownames(hallmark.GSEA.wide.m) <- hallmark.GSEA.wide$pathway
#hallmark.GSEA.wide.m <- hallmark.GSEA.wide.m[rowSums(is.na(hallmark.GSEA.wide.m)) < 4, ] # keeping only pathways that are not NA in more than 3 samples
```


```{r}
quantile(as.matrix(hallmark.GSEA.wide.m), 
         c(0.1, 0.9), na.rm = TRUE) # doing this to determine relevant min and max value to cap colour scale for heat map
```

```{r}
library(circlize)
col_fun <- colorRamp2(c(-1.8, 0, 1.8), viridis(3))
```


```{r}
ht_list <- Heatmap(hallmark.GSEA.wide.m, 
        name = "NES, GSEA", 
        na_col = "black", 
        column_title = "Comparisons", 
    row_title = "Pathways", 
    column_title_side = "bottom", 
    column_title_gp = gpar(fontsize = 8, fontface = "bold"), 
    row_title_gp = gpar(fontsize = 8, fontface = "bold"), 
    column_names_gp = gpar(fontsize = 8),
    row_names_gp = gpar(fontsize = 5),
    border = TRUE,
    cluster_rows = TRUE, # must be FALSE if filtering on padj after rbind() above
    cluster_columns = TRUE, # must be FALSE if filtering on padj after rbind() above
    col = col_fun,
    use_raster = TRUE, 
    raster_device = "png",
    show_row_names = TRUE
    )
pdf(paste0(output.figures, Sys.Date(), 
           sep="_", "heatmap_GSEA_from_t-values", ".pdf"), width = 5, height = 5)
draw(ht_list, heatmap_legend_side = "left")
dev.off()
```


# PROGENy

Alternative pathway analysis method based on Schubert et al. 2018 Nature Communications (doi: 10.1186/s13059-020-1949-z)

- Need voom matrix with HGNC identifiers

```{r}
v.DGE.df <- as.data.frame(v.DGE) %>% 
  mutate(Genes = rownames(v.DGE))
```

```{r}
v.DGE.annot <-
  inner_join(v.DGE.df, de.data.annot.no.dupl, by = "Genes") # inner join to make sure only Genes without HGNC duplications or NA retained
v.DGE.annot 
```

```{r}
colnames(v.DGE.annot)
```


```{r}
v.DGE.annot.m <- v.DGE.annot %>% 
  dplyr::select(contains("HGNC") | contains("WT_vehicle") | contains("KO_vehicle") |
                  contains("WT_inh") | contains("KO_inh")) %>% 
  dplyr::select(-c(HGNC, contains("P.value") | contains("Coef.") | contains("t."))) %>% 
  dplyr::mutate_if(~ any(is.na(.x)),~ if_else(is.na(.x),0,.x)) %>% 
  as.matrix() 
rownames(v.DGE.annot.m) <- v.DGE.annot %>% 
  .$HGNC
```

```{r}
head(v.DGE.annot.m)
```


```{r}
progeny <- progeny(v.DGE.annot.m, scale=TRUE, 
    organism="Human", top = 100)
Activity_counts <- as.vector(progeny)
```


```{r}
paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(Activity_counts), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(Activity_counts)/paletteLength, 
    max(Activity_counts), 
    length.out=floor(paletteLength/2)))

progeny_hmap <- pheatmap(t(progeny),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    border_color = NA)
```

```{r}
quantile(progeny, 
         c(0.1, 0.9), na.rm = TRUE) # doing this to determine relevant min and max value to cap colour scale for heat map
```

```{r}
col_fun <- colorRamp2(c(-1.3, 0, 1.3), viridis(3))
```


```{r}
ht_list <- Heatmap(t(progeny), 
        name = "PROGENy Score", 
        na_col = "black", 
        column_title = "Samples", 
    row_title = "Pathways", 
    column_title_side = "bottom", 
    column_title_gp = gpar(fontsize = 8, fontface = "bold"), 
    row_title_gp = gpar(fontsize = 8, fontface = "bold"), 
    column_names_gp = gpar(fontsize = 8),
    row_names_gp = gpar(fontsize = 5),
    border = TRUE,
    cluster_rows = TRUE, # must be FALSE if filtering on padj after rbind() above
    cluster_columns = TRUE, # must be FALSE if filtering on padj after rbind() above
    col = col_fun,
    use_raster = TRUE, 
    raster_device = "png",
    show_row_names = TRUE
    )
pdf(paste0(output.figures, Sys.Date(), 
           sep="_", "heatmap_PROGENy", ".pdf"), width = 5, height = 5)
draw(ht_list, heatmap_legend_side = "left")
dev.off()
```


***

